---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Mason Ibrahim"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
#Verify working directory
getwd()

#Load in packages
library(tidyverse)
library(agricolae)   
library(here)
library(ggthemes)
here()

#Read in data
NTL_raw <- 
  read.csv(here
           ("~/R/EDE_Fall2024/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
                    , stringsAsFactors = TRUE)

#set dates
NTL_raw$sampledate <- as.Date(NTL_raw$sampledate, format = "%m/%d/%y")
```


```{r theme}
#2
my_theme <- theme_base() +
  theme(
    rect = element_rect(color = "darkblue"),
    text = element_text(color = "blue"),
    plot.title = element_text(
      size = 16,
      face = "bold",
      color = "blue"
    ),
    axis.title.x = element_text(size = 12, color = "blue"),
    axis.title.y = element_text(size = 12, color = "blue"),
    axis.text = element_text(size = 12, color = "lightblue"),
    
    axis.ticks = element_line(color = "lightblue"),
    panel.grid.major = element_line(color = "lightblue"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = "blue"),
    legend.background = element_rect(fill = "lightblue", color = "blue")
  )

complete = TRUE
theme_set(my_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature from July does not change with depth across all lakes.
Ha: Mean lake temperature from July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 째C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_wrangled<-
  NTL_raw %>% 
  filter(month(sampledate) == 7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  na.omit()
  

#5
NTL_wrangle_plot <-
  ggplot(NTL_wrangled, aes(x = depth, y = temperature_C))+
  geom_point(alpha = .25,
             col = "lightblue3") +
  geom_smooth(method = "lm",
              col = "darkblue")+
  ylim(0, 35)+
  labs(x = "Depth (meters)", y = "Temperature (째C)")
print(NTL_wrangle_plot)


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It seems that temperature decreases as depth increases. However, the distribution of the points suggests that this linearity ends at around 7 meters, where then the points start to level out around 5-7 degress celcius.There also seems to be a high amount of variability in temperatures at these depths.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
lm_NTL_temp_depth <- lm(
  data = NTL_wrangled,
  temperature_C ~ depth)
summary(lm_NTL_temp_depth)

par(mfrow = c(2,2), mar=c(4,4,4,4))
plot(lm_NTL_temp_depth)
par(mfrow = c(1,1))

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model shows that depth significantly affects temperature. For every 1-meter increase in depth, temperature is predicted to decrease by about 1.95째C. Depth explains 73.87% of the variability in temperature. This finding is based on 9726 degrees of freedom and is statistically significant with a p-value of <2.2e-16, which is less than .05.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
NTL_aic <- lm(data = NTL_wrangled , temperature_C ~ depth + year4 + daynum)
step(NTL_aic)

#10
summary(NTL_aic)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests we use to predict temperature in the multiple regression are depth, year4, and daynum. This model explains 74.12% of the observed variance. This is a small improvement over the model that only used depth as the explanatory variable which explained 73.87%.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
NTL_anova <- aov(data = NTL_wrangled, temperature_C ~ lakename)
summary(NTL_anova)

NTL_anova_lm <- lm(data = NTL_wrangled, temperature_C ~ lakename)
summary(NTL_anova_lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference in mean temperature among the lakes with a p-value less that 0.05. However, it explains only about 3.95% of the variance in temperature. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
NTL_lm_plot <-
  ggplot(NTL_wrangled, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = .5) +
  geom_smooth(method = "lm",
              se = FALSE)+
  ylim(0, 35)+
  labs(x = "Depth (meters)", y = "Temperature (째C)", color = "Lake Name")
print(NTL_lm_plot)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(NTL_anova)
NTL_names_group <- HSD.test(NTL_anova, "lakename", group = TRUE)
NTL_names_group
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and Ward Lake have similar mean temperature as Peter Lake. No lake is statistically distinct from all the other lakes.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two-sample t-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL_crampton_ward <-
  NTL_wrangled %>% 
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

NTL_crampton_ward_ttest <- 
  t.test(temperature_C ~ lakename, data = NTL_crampton_ward, var.equal = TRUE)
print(NTL_crampton_ward_ttest)
```

>Answer: The mean temperatures are not equal. The result is Crampton Lake's mean is 15.35 and Ward Lake's mean is 14.46. However, the p-value is 0.2592, which is not less that 0.05, meaning the difference is not statistacally significant.
This does match what I got from the Tukey's Test, in which Crampton Lake was in group ab and Ward was in group bc, meaning they were statistically similar to each other because they both have b.

